---
title: "differential_expression_analysis"
author: "Mohit Rastogi"
date: '`r format(Sys.Date(), "%B %d, %Y")`'
output: workflowr::wflow_html
editor_options:
  chunk_output_type: console
---



## Introduction

This is the differential expression analysis from the total RNA sequencing experiment performed on postmortem hippocampus obtained from Down syndrome and control individuals.


#### First, we will load all the packages required to perform this analysis
```{r setup, message = FALSE, warning=FALSE,include=TRUE}
library(ggplot2)
library(dplyr)
library(edgeR)
library(openxlsx)
library(ggrepel)
library(DBI)
library(org.Hs.eg.db)
library("pcaExplorer")
library(TxDb.Hsapiens.UCSC.hg38.knownGene)
library(rtracklayer)
library(ggbio)
library(clusterProfiler)
library(enrichplot)
library(ggridges)
library(karyoploteR)
library(tidyverse)
library(forcats)
library(pathview)
library(ComplexHeatmap)
library(circlize)
library(kableExtra)
library(DT)
```

#### Read the raw count matrix for hippocampus samples

The count matrix consists of 11 control samples (C1-C11) and 9 Down syndrome samples (Ds1-Ds9)

```{r, echo = TRUE, warning=FALSE,include=TRUE}
count_human_total_hippocampus_coding<- read.table(file= "data/raw_counts_human_hippo_matrix_coding.txt", header=T, row.names=1)
head(count_human_total_hippocampus_coding)
meta_counts_human_total_hippocampus_coding<-data.frame(row.names = colnames(count_human_total_hippocampus_coding),condition_human_total_hippocampus_coding=c("Cont","Cont","Cont","Cont","Cont","Cont","Cont","Cont","Cont","Cont","Cont","DS","DS","DS","DS","DS","DS","DS","DS","DS"))
```

Define the groups

```{r, echo = TRUE, warning=FALSE,include=TRUE}
group_human_total_hippocampus_coding<-relevel(factor(meta_counts_human_total_hippocampus_coding$condition_human_total_hippocampus_coding),ref="Cont")
group_human_total_hippocampus_coding
```

Design the model for performing differential expression analysis

```{r, echo = TRUE, warning=FALSE,include=TRUE}
#Design the model for the differential expression
design_human_total_hippocampus_coding = model.matrix(~group_human_total_hippocampus_coding)
y_human_total_hippocampus_coding<-DGEList(count_human_total_hippocampus_coding,group = group_human_total_hippocampus_coding)
```

Calculate counts per million in log

```{r, echo = TRUE, warning=FALSE,include=TRUE}
cpm_log_human_total_hippocampus_coding<-cpm(y_human_total_hippocampus_coding,log=TRUE)
```

#### Visualize the log counts per million (cpm) as a heatmap
```{r, echo = TRUE, warning=FALSE,include=TRUE}
heatmap(cor(cpm_log_human_total_hippocampus_coding))
```

#### Perform PCA after loading the sample information
```{r, echo = TRUE, warning=FALSE,include=TRUE}
sample_hippo<- read.xlsx("data/human_hippo_samplesheet.xlsx", sheet=1)
row.names(sample_hippo)<-sample_hippo$SampleName
all(colnames(cpm_log_human_total_hippocampus_coding) == rownames(sample_hippo))
p <- PCAtools::pca(cpm_log_human_total_hippocampus_coding, metadata = sample_hippo, removeVar = 0.1)
PCAtools::screeplot(p)
PCAtools::biplot(p)
```

#### Filter the data based on heatmap, PCA and RIN values

Looking at the clustering and PCA plots, we remove few samples which look like outliers: C6,C7,C8,C9,C10,Ds2,Ds6,Ds7

```{r, echo = TRUE, warning=FALSE,include=TRUE}
count_hth_filt_coding2<-count_human_total_hippocampus_coding[,-c(6,7,8,9,10,13,17,18)]
head(count_hth_filt_coding2)
meta_counts_hth_filt_coding2<-data.frame(row.names = colnames(count_hth_filt_coding2),condition_hth_filt_coding2=c("Cont","Cont","Cont","Cont","Cont","Cont","DS","DS","DS","DS","DS","DS"))
```

####Perform hierarchical clsutering and PCA on the filtered samples

```{r, echo = TRUE, warning=FALSE,include=TRUE}
group_hth_filt_coding2<-relevel(factor(meta_counts_hth_filt_coding2$condition_hth_filt_coding2),ref="Cont")
group_hth_filt_coding2
design_hth_filt_coding2 = model.matrix(~group_hth_filt_coding2)
y_hth_filt_coding2<-DGEList(count_hth_filt_coding2,group = group_hth_filt_coding2)
cpm_log_hth_filt_coding2<-cpm(y_hth_filt_coding2,log=TRUE)
heatmap(cor(cpm_log_hth_filt_coding2))
sample_hippo<- read.xlsx("data/human_hippo_samplesheet.xlsx", sheet=2)
row.names(sample_hippo)<-sample_hippo$SampleName
all(colnames(cpm_log_hth_filt_coding2) == rownames(sample_hippo))
p <- PCAtools::pca(cpm_log_hth_filt_coding2, metadata = sample_hippo, removeVar = 0.1)
PCAtools::screeplot(p)
PCAtools::biplot(p)
```

#### Filter the genes based on their expression in n-1 samples
```{r, echo = TRUE, warning=FALSE,include=TRUE}
cpm_y_hth_filt_coding2<-cpm(y_hth_filt_coding2)
keep_hth_filt_coding2<-rowSums((cpm_y_hth_filt_coding2)>1)>5
table(keep_hth_filt_coding2)
```

#### Perform normalization using TMM
```{r, echo = TRUE, warning=FALSE,include=TRUE}
y_hth_filt_coding2<-DGEList(count_hth_filt_coding2,group = group_hth_filt_coding2)
y_hth_filt_coding2 <- y_hth_filt_coding2[keep_hth_filt_coding2, , keep.lib.sizes=FALSE]
y_hth_filt_coding2 <- calcNormFactors(y_hth_filt_coding2)
y_hth_filt_coding2$samples
```

#### Estimate differential expression using GLM
```{r, echo = TRUE, warning=FALSE,include=TRUE}
y_hth_filt_coding2 <- estimateGLMRobustDisp(y_hth_filt_coding2,design_hth_filt_coding2, verbose = TRUE)
fit_hth_filt_coding2 <- glmFit(y_hth_filt_coding2, design_hth_filt_coding2)
lrt_hth_filt_coding2<-glmLRT(fit_hth_filt_coding2,coef = 2)
de_hth_filt_coding2 <- decideTestsDGE(lrt_hth_filt_coding2, adjust.method="BH", p.value = 0.05)
```

#### Number of differentially expressed genes in the hippocampus
```{r, echo = TRUE, warning=FALSE,include=TRUE}
summary(de_hth_filt_coding2)
```

#### Counts per million (CPM) values for all the genes considered in the differential expression analysis

```{r, echo = TRUE, warning=FALSE,include=TRUE}
cpm_y_hth_filt_coding22<-cpm(y_hth_filt_coding2)
cpm_y_hth_filt.dt <- DT::datatable(cpm_y_hth_filt_coding22, rownames=TRUE, class="white-space: nowrap", escape=FALSE) 
cpm_y_hth_filt.dt
```


#### Here, we annotated the data from the previous using ENSEMBL information for each gene and mapped the DEGs onto their chromosomal location.

```{r, echo = TRUE, warning=FALSE,include=TRUE}
OrgDb <- org.Hs.eg.db
results_edgeR_hth_filt_coding2<- topTags(lrt_hth_filt_coding2, n = nrow(count_hth_filt_coding2), sort.by = "none")
k<-row.names(results_edgeR_hth_filt_coding2)
ann_hth_filt_coding2<-AnnotationDbi::select(org.Hs.eg.db,keys=k,keytype = "ENSEMBL",columns=c("ENTREZID","SYMBOL","GENENAME","CHR","UNIPROT","ALIAS","GENENAME"))
idx<-match(row.names(results_edgeR_hth_filt_coding2),ann_hth_filt_coding2$ENSEMBL)
results_rna_annotated_hth_filt_coding2<-cbind(results_edgeR_hth_filt_coding2,ann_hth_filt_coding2[idx,])
detags <- rownames(y_hth_filt_coding2)[as.logical(de_hth_filt_coding2)]
sigGenes <- results_rna_annotated_hth_filt_coding2[detags,]
tx<-TxDb.Hsapiens.UCSC.hg38.knownGene
exo <- exonsBy(tx,"gene")
exoRanges <- unlist(range(exo))
sigRegions <- exoRanges[na.omit(match(sigGenes$ENTREZID, names(exoRanges)))]
mcols(sigRegions) <- sigGenes[match(names(sigRegions), sigGenes$ENTREZID),]
sigRegions[order(sigRegions$LR,decreasing = TRUE)]
sigRegions <- keepSeqlevels(sigRegions, value = c("chr1","chr2","chr3","chr4","chr5","chr6","chr7",
"chr8","chr9","chr10","chr11","chr12","chr13","chr14",
"chr15","chr16","chr17","chr18","chr19","chr20",
"chr21","chr22","chrX","chrY"),pruning.mode="tidy")
#seqlevels(sigRegions)
Score <- -log10(sigRegions$FDR)
rbPal <-colorRampPalette(c("blue", "red"))
logfc <- pmax(sigRegions$logFC, -5)
logfc <- pmin(logfc , 5)
Col <- rbPal(10)[as.numeric(cut(logfc, breaks = 10))]
mcols(sigRegions)$score <- Score
mcols(sigRegions)$itemRgb <- Col
#export(sigRegions , con = "topHitshuman_hippo.bed")
top200 <- sigRegions[order(sigRegions$LR,decreasing = TRUE)]
plotGrandLinear(top200 , aes(y = logFC))
mcols(top200)$UpRegulated <- mcols(top200)$logFC > 0
mcols(top200)$DownRegulated <- mcols(top200)$logFC < 0
autoplot(top200,layout="karyogram",aes(color=DownRegulated,fill=DownRegulated))
results_hippo_RNA<-as_tibble(results_rna_annotated_hth_filt_coding2)
row.names(results_hippo_RNA)<-results_hippo_RNA$ENSEMBL
saveRDS(results_hippo_RNA, file = "output/results_hippo_RNA.rds")
saveRDS(sigGenes,file="output/sigGenes_hippo.rds")
```

#### Results from differential expression analysis for hippocampus samples

```{r, echo = TRUE, warning=FALSE,include=TRUE}
results_rna_annotated_hth_filt_coding2.dt <- DT::datatable(results_rna_annotated_hth_filt_coding2, rownames=TRUE, class="white-space: nowrap", escape=FALSE) 
results_rna_annotated_hth_filt_coding2.dt
```
